\section{History of Machine Learning Libraries}\label{sec:history}

In this section, we aim to give a brief overview and key milestones in the
history of machine learning software libraries. We begin with a review of
libraries suitable for a wide range of machine learning and data analysis
purposes, reaching back more than 20 years. We then perform a more focused study
of recent programming frameworks suited especially to the task of deep
learning. Figure \ref{fig:timeline} visualizes this section in a timeline. We
wish to emphasize that this section does in no way compare TensorFlow, as we
have dedicated Section \ref{sec:comp} to this specific purpose.

\subsection{General Machine Learning}\label{sec:history-general}

In the following paragraphs we list and briefly review a small set of
\emph{general machine learning libraries} in chronological order. With
\emph{general}, we mean to describe any particular library whose common
use cases in the machine learning and data science community include \emph{but
  are not limited to} deep learning. As such, these libraries may be used for
statistical analysis, clustering, dimensionality reduction, structured
prediction, anomaly detection, shallow (as opposed to deep) neural networks and
other tasks.

We begin our review with a library published 21 years before TensorFlow:
\emph{MLC++} \cite{mlcpp}. MLC++ is a software library developed in the C++
programming language providing algorithms alongside a comparison framework for a
number of data mining, statistical analysis as well as pattern recognition
techniques. It was originally developed at Stanford University in 1994 and is
now owned and maintained by Silicon Graphics, Inc
(SGI\footnote{https://www.sgi.com/tech/mlc/}). To the best of our knowledge,
MLC++ is the oldest machine learning library still available today.

Following MLC++ in the chronological order,
\emph{OpenCV}\footnote{http://opencv.org} (\textbf{O}pen
\textbf{C}omputer \textbf{V}ision) was released in the year 2000 by Bradski et
al. \cite{opencv}. It is aimed primarily at solving learning tasks in the field
of computer vision and image recognition, including a collection of algorithms
for face recognition, object identification, 3D-model extraction and other
purposes. It is released under a BSD license and provides interfaces in multiple
programming languages such as C++, Python and MATLAB.

Another machine learning library we wish to mention is
\emph{scikit-learn}\footnote{http://scikit-learn.org/stable/} \cite{scikit}. The
scikit-learn project was originally developed by David Cournapeu as part of the
Google Summer of Code program\footnote{https://summerofcode.withgoogle.com} in
2008. It is an open source machine learning library written in Python, on top of
the NumPy, SciPy and matplotlib frameworks. It is useful for a large class of
both supervised and unsupervised learning problems.

The \emph{Accord.NET}\footnote{http://accord-framework.net/index.html} library
stands apart from the aforementioned examples in that it is written in the C\#
(``C Sharp'') programming language. Released in 2008, it is composed not only of
a variety of machine learning algorithms, but also signal processing modules for
speech and image recognition \cite{accord}.

\emph{Massive Online Analysis}\footnote{http://moa.cms.waikato.ac.nz} (MOA) is
an open source framework for online and offline analysis of massive, potentially
infinite, data \emph{streams}. MOA includes a variety of tools for
classification, regression, recommender systems and other disciplines. It is
written in the Java programming language and maintained by staff of the
University of Waikato, New Zealand. It was conceived in 2010 \cite{moa}.

The \emph{Mahout}\footnote{http://mahout.apache.org} project, part of Apache
Software Foundation\footnote{http://www.apache.org}, is a Java programming
environment for scalable machine learning applications, built on top of the
Apache Hadoop\footnote{http://hadoop.apache.org} platform. It allows for
analysis of large datasets distributed in the Hadoop Distributed File System
(HDFS) using the \emph{MapReduce} programming paradigm. Mahout provides
machine learning algorithms for classification, clustering and filtering.

\emph{Pattern}\footnote{http://www.clips.ua.ac.be/pages/pattern} is a Python
machine learning module we include in our list due to its rich set of
\emph{web mining} facilities. It comprises not only general machine learning
algorithms (e.g. clustering, classification or nearest neighbor search) and
natural language processing methods (e.g. n-gram search or sentiment analysis),
but also a web crawler that can, for example, fetch Tweets or Wikipedia entries,
facilitating quick data analysis on these sources. It was published by the
University of Antwerp in 2012 and is open source.

Lastly, \emph{Spark MLlib}\footnote{http://spark.apache.org/mllib} is an
open source machine learning and data analysis platform released in 2015 and
built on top of the Apache Spark\footnote{http://spark.apache.org/} project
\cite{spark}, a fast cluster computing system. Similar to Apache Mahout, it
supports processing of large scale \emph{distributed} datasets and training of
machine learning models across a cluster of commodity hardware. For this, it
includes classification, regression, clustering and other machine learning
algorithms \cite{mllib}.

\begin{figure*}[t!]
\centering
  \begin{tikzpicture}
  \draw [thick, ->] (0, 0) -- (13, 0);

  \newcounter{y}
  \setcounter{y}{1992}
  \foreach \i in {0, ..., 12} {
    \draw (\i, {(Mod(\i, 2) - 0.5)/1.5}) node {\they};
    %\draw (\i, -0.1) -- (\i, +0.1);
    \stepcounter{y}
    \stepcounter{y}
  }

  \foreach \time/\lib/\a/\b/\c in {%
    1994/MLC++/-1/0/red,
    2000/OpenCV/+1/0/blue,
    2002/Torch/-1/0/teal,
    2007/scikit/+1/0/magenta,
    2008/Accord/-1/+0.1/orange,
    2008/Theano/-1.5/-0.1/black,
    2010/MOA/+1/0/cyan,
    2011/Mahout/-1/0/teal,
    2012/Pattern/+1.7/-0.1/red,
    2014/DL4J/-0.5/0.15/blue,
    2014/Caffe/+1/-0.15/olive,
    2014/cuDNN/-1.5/0/magenta,
    2015/TensorFlow/+1.5/0/orange,
    2015/MLlib/-1/0.2/black%
    } {
    \draw [\c] ({(\time - 1992)/2 + \b}, \a) node [align=center] {\lib};
    \filldraw [\c] ({(\time - 1992)/2 + \b}, 0) circle [radius=1.5pt];
  }

  \end{tikzpicture}
\caption{A timeline showing the release of machine-learning libraries discussed
  in section I in the last 25 years.}
\label{fig:timeline}
\end{figure*}

\subsection{Deep Learning}\label{sec:history-dl}

While the software libraries mentioned in the previous section are useful for a
great variety of different machine learning and statistical analysis tasks, the
following paragraphs list software frameworks especially effective in training
deep learning models.

The first and oldest framework in our list suited to the development and
training of deep neural networks is \emph{Torch}\footnote{http://torch.ch},
released already in 2002 \cite{torch}. Torch consisted originally of a pure C++
implementation and interface. Today, its core is implemented in C/CUDA while it
exposes an interface in the Lua\footnote{https://www.lua.org} scripting
language. For this, Torch makes use of a LuaJIT (just-in-time) compiler to
connect Lua routines to the underlying C implementations. It includes, inter
alia, numerical optimization routines, neural network models as well as
general purpose n-dimensional array (tensor) objects.

\emph{Theano}\footnote{http://deeplearning.net/software/theano/}, released in
2008 \cite{theano}, is another noteworthy deep learning library. We note that
while Theano enjoys greatest popularity among the machine learning community, it
is, in essence, not a machine learning library at all. Rather, it is a
programming framework that allows users to declare mathematical expressions
\emph{symbolically}, as computational graphs. These are then optimized,
eventually compiled and finally executed on either CPU or GPU devices. As such,
\cite{theano} labels Theano a ``mathematical compiler''.

\emph{Caffe}\footnote{http://caffe.berkeleyvision.org} is an open source
deep learning library maintained by the Berkeley Vision and Learning Center
(BVLC). It was released in 2014 under a BSD-License \cite{caffe}. Caffe is
implemented in C++ and uses neural network layers as its basic computational
building blocks (as opposed to Theano and others, where the user must define
individual mathematical operations making up layers). A deep learning model,
consisting of many such layers, is stored in the Google Protocol Buffer
format. While models can be defined manually in this Protocol Buffer
``language'', there exist bindings to Python and MATLAB to generate them
programmatically. Caffe is especially well suited to the development and
training of \emph{convolutional neural networks} (CNNs or ConvNets), used
extensively in the domain of image recognition.

While the aforementioned machine learning frameworks allowed for the definition
of deep learning models in Python, MATLAB and Lua, the
\emph{Deeplearning4J}\footnote{http://deeplearning4j.org} (DL4J) library enables
also the Java programmer to create deep neural networks. DL4J includes
functionality to create Restricted Boltzmann machines, convolutional and
recurrent neural networks, deep belief networks and other types of deep learning
models. Moreover, DL4J enables horizontal scalability using distributed
computing platforms such as Apache Hadoop or Spark. It was released in 2014 by
Adam Gibson under an Apache 2.0 open source license.

Lastly, we add the NVIDIA Deep Learning
SDK\footnote{https://developer.nvidia.com/deep-learning-software} to to this
list. Its main goal is to maximize the performance of deep learning algorithms
on (NVIDIA) GPUs. The SDK consists of three core modules. The first,
\emph{cuDNN}, provides high performance GPU implementations for deep learning
algorithms such as convolutions, activation functions and tensor
transformations. The second is a linear algebra library, \emph{cuBLAS}, enabling
GPU-accelerated mathematical operations on n-dimensional arrays. Lastly,
\emph{cuSPARSE} includes a set of routines for \emph{sparse} matrices tuned for
high efficiency on GPUs. While it is possible to program in these libraries
directly, there exist also bindings to other deep learning libraries, such as
Torch\footnote{https://github.com/soumith/cudnn.torch}.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../paper"
%%% End:
